---
title: "36-315 Final Project"
author: "Kyra Balenzano, Evan Feder, David Yuan"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r message = F, warning = F}
spotify <- read.csv("spotify-2000.csv") # Load the data

# Load the required libraries
library(tidyverse)
library(pander)
```

# DATA DESCRIPTION

```{r}
# EDA

library(gridExtra)

year.hist = ggplot(spotify, aes(x = Year)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
bpm.hist = ggplot(spotify, aes(x = `Beats Per Minute (BPM)`)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
energy.hist = ggplot(spotify, aes(x = Energy)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
dance.hist = ggplot(spotify, aes(x = Danceability)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
loud.hist = ggplot(spotify, aes(x = `Loudness (dB)`)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
live.hist = ggplot(spotify, aes(x = Liveness)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
val.hist = ggplot(spotify, aes(x = Valence)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
len.hist = ggplot(spotify, aes(x = `Length (Duration)`)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
acou.hist = ggplot(spotify, aes(x = Acousticness)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
speech.hist = ggplot(spotify, aes(x = Speechiness)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
pop.hist = ggplot(spotify, aes(x = Popularity)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)

grid.arrange(year.hist, bpm.hist, energy.hist, dance.hist, loud.hist, live.hist,
             val.hist, len.hist, acou.hist, speech.hist, pop.hist, nrow = 4)
```


***
***

# RESEARCH QUESTIONS

+ What attributes are associated with which genres?
  - Facetted histograms
  - Rap songs have more speechiness
  - Few songs have high liveness or acousticness
  - 
+ What attributes make a song popular?
  - Plot danceability vs. popularity
+ Have the attributes of songs that appear on the top 2000 list changed over time?
  - MDS plot/dendrogram, comparison word cloud for top genre


***
***


# GRAPHS

## Question 1

### Graph 1

### Graph 2

### Graph 3


***


## Question 2

### Graph 1

### Graph 2

### Graph 3


***


## Question 3

### Graph 1

### Graph 2

### Graph 3


***
***


# CONCLUSION

