---
title: "36-315 Final Project"
author: "Kyra Balenzano, Evan Feder, David Yuan"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r message = F, warning = F}
spotify <- read.csv("spotify-2000.csv") # Load the data

# Easier column names
colnames(spotify)[1] <- "Index"
colnames(spotify)[6] <- "BPM"
colnames(spotify)[9] <- "Loudness"
colnames(spotify)[12] <- "Duration"

# Load the required libraries
library(tidyverse)
library(pander)
library(gridExtra)
```

# DATA DESCRIPTION

```{r}
# Cleaning up the genres column
rock <- c("album rock", "alternative metal", "classic rock", "modern rock", 
          "alternative rock", "garage rock", "permanent wave", "modern folk rock",
          "irish rock", "art rock", "celtic rock", "dutch rock", "belgian rock",
          "british invasion", "finnish metal", "dutch metal", "soft rock", "dutch prog",
          "dance rock", "mellow gold", "glam metal", "australian rock", "australian psych",
          "rock-and-roll", "glam rock", "hard rock", "punk", "j-core", "australian
          alternative rock", "yacht rock", "celtic punk", "classic canadian rock",
          "cyberpunk", "classical rock", "christelijk", "canadian rock", "british
          alternative rock", "german alternative rock")
pop <- c("alternative pop rock", "pop", "classic uk pop", "dance pop", "dutch pop",
         "alternative dance", "german pop", "afropop", "disco", "danish pop rock",
         "britpop", "neo mellow", "boy band", "hip pop", "australian pop", "canadian pop",
         "bow pop", "acoustic pop", "candy pop", "operatic pop", "alternative pop",
         "eurodance", "art pop", "uk pop", "brill building pop", "belgian pop",
         "barbadian pop", "chamber pop", "indie pop", "electropop", "folk-pop",
         "metropopolis", "irish pop", "australian dance", "nederpop", "danish pop",
         "italian pop", "la pop", "baroque pop", "austropop", "ccm", "bubblegum pop",
         "europop", "new wave pop", "german pop rock", "levenslied", "classic 
         italian pop", "pop punk")
country <- c("dutch americana", "arkansas country", "british folk", "blues rock",
             "canadian folk", "contemporary country", "australian indie folk", 
             "indie anthem-folk", "australian americana", "classic country pop",
             "folk", "alternative country")
indie <- c("dutch indie", "alaska indie", "icelandic indie")
hip_hop <- c("alternative hip top", "detroit hip hop", "east coast hip hop", 
             "dutch hip hop", "atl hip hop", "g funk", "gangster rap")

spotify <- spotify %>% 
  mutate(Genre = ifelse(Top.Genre %in% rock, "rock",
                 ifelse(Top.Genre %in% pop, "pop",
                 ifelse(Top.Genre %in% country, "country",
                 ifelse(Top.Genre %in% indie, "indie",
                 ifelse(Top.Genre %in% hip_hop, "hiphop",
                        "other")))))) %>% 
  mutate(Genre = factor(Genre),
         Duration = as.integer(Duration)) %>% 
  select(-Top.Genre)
```

```{r}
# EDA
year.hist = ggplot(spotify, aes(x = Year)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
bpm.hist = ggplot(spotify, aes(x = BPM)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
energy.hist = ggplot(spotify, aes(x = Energy)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
dance.hist = ggplot(spotify, aes(x = Danceability)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
loud.hist = ggplot(spotify, aes(x = Loudness)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
live.hist = ggplot(spotify, aes(x = Liveness)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
val.hist = ggplot(spotify, aes(x = Valence)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
len.hist = ggplot(spotify, aes(x = Duration)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
acou.hist = ggplot(spotify, aes(x = Acousticness)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
speech.hist = ggplot(spotify, aes(x = Speechiness)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
pop.hist = ggplot(spotify, aes(x = Popularity)) + 
  geom_histogram(fill = "lightblue", color = "black", bins = 20)
genre.bar = ggplot(spotify, aes(x = Genre, fill = Genre)) +
  geom_bar(color = "black")

grid.arrange(year.hist, bpm.hist, energy.hist, dance.hist, loud.hist, live.hist,
             val.hist, len.hist, acou.hist, speech.hist, pop.hist, genre.bar, 
             nrow = 3)
```


***
***

# RESEARCH QUESTIONS

+ What attributes are associated with which genres?
  - Facetted histograms
  - Rap songs have more speechiness
  - Few songs have high liveness or acousticness
  - 
+ What attributes make a song popular?
  - Plot danceability vs. popularity
+ Have the attributes of songs that appear on the top 2000 list changed over time?
  - MDS plot/dendrogram, comparison word cloud for top genre


***
***


# GRAPHS

## Question 1

### Graph 1

### Graph 2

### Graph 3


***


## Question 2

### Graph 1

### Graph 2

### Graph 3


***


## Question 3

### Graph 1

### Graph 2

### Graph 3


***
***


# CONCLUSION

